---
title: "futurePA - Future vertebrate richness changes in protected areas under climate change"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(collapse = T, comment = NA, warning=F, message=F, echo=F, 
                      fig.path="figures/", dpi=600)
```

<!--
## MATERIAL AND METHODS

Global protected area data was downloaded from protectedplanet.net in Feburary 2018.  

We consider temperature, precipitation, seasonality in temperature and precipitation, elevation, isolation, and connectivity. Using bioclimatic data for current and future conditions and IUCN range maps, we model current and future global vertebrate richness (amphibians, mammals and birds) of current protected areas to assess its coverage with regard to global biodiversity.

Species-specific IUCN range maps were used as response variable of the SDMs and to create current species richness maps for each of the three taxonomic groups. Bioclimatic variables derived from the meteorological forcing dataset EartH2Observe, WFDEI and ERA-Interim data Merged and Bias-corrected for ISIMIP (EWEMBI) were used as explanatory variables for the SDMs. We compare species richness of protected areas with specis richness of non-protected areas. We then create future predictions using future bioclimatic variables of 4 global climate models (GFDL-ESM2M, MIROC5, HadGEM2-ES, IPSL-CM5A-LR) and two representative concentration pathways (RCP2.6 and RCP6.0) and compare the future predictions of species richness among the protected and non-protected areas.

## RESULTS

As of August 2014, there are currently 209,429 designated protected areas worldwide, covering an area of 32,868,673 km2 (Deguignet et al. 2014).

We obtained polygon data for 217825 (203167 terrestrial and 14658 coastal & marine) protected areas and point data for 18406 (17460 terrestrial and 946 coastal & marine) protected areas.
-->

## Maps of PAs

```{r pas_1995, fig.width=10, fig.height=5}
#' Load required libraries
library(sf); library(raster); library(tidyverse)

#' Load Github package
library(rISIMIP)

########################################

# Load protected area 0.5deg data
pas <- get(data("protectedareas_annual_1819_2018_landonly", package="rISIMIP"))
#colnames(pas)
rm(protectedareas_annual_1819_2018_landonly); invisible(gc())

# Plot map of PAs with more than 0.5
library(rnaturalearthhires)
data(coastline10)
coastline10 <- sf::st_as_sf(coastline10)

pas %>% filter(`1995` >= 0.5) %>% 
  ggplot() + geom_tile(aes(x,y,fill=`1995`)) + theme_bw() + 
  scale_fill_gradientn(colours = rev(terrain.colors(10))) + 
  geom_sf(data=coastline10, colour="black", fill=NA) + 
  coord_sf() + labs(x="", y="")
#ggsave("figures/pas_1995.png", dpi=600, width=10, height=5)
```

**Fig. 1.** Protected areas and their percentage coverage in 1995.

```{r pas_2018,  fig.width=10, fig.height=5}
pas %>% filter(`2018` >= 0.5) %>% 
  ggplot() + geom_tile(aes(x,y,fill=`2018`)) + theme_bw() + 
  scale_fill_gradientn(colours = rev(terrain.colors(10))) + 
  geom_sf(data=coastline10, colour="black", fill=NA) + 
  coord_sf() + labs(x="", y="")
#ggsave("figures/pas_2018.png", dpi=600, width=10, height=5)
```
**Fig. 2.** Protected areas and their percentage coverage in 2018.

```{r bio1_pas}
########################################

#' ## Environmental data

#' Get bioclim files
datafiles <- data(package="rISIMIP")
datafiles <- as.data.frame(datafiles$results)
climfiles <- datafiles[grep(datafiles$Item, pattern="bioclim"),]

#' Read ISIMIP bioclim data as list
climatedata <- lapply(sub('\"', '', sub('','"', climfiles$Item)), function(x) get(data(list=x, package="rISIMIP")))
names(climatedata) <- sub('\"', '', sub('','"', climfiles$Item))
ewembi <- climatedata[[1]]
climatedata <- dplyr::bind_rows(climatedata[2:9], .id="id")
rm(list=ls(pattern="bioclim")); invisible(gc())

#' Add model, gcm, rcp to data
ewembi$gcm <- "EWEMBI"
ewembi$rcp <- ""
ewembi$year <- 1995
climatedata$id <- sub("bioclim_", "", sub("_landonly", "", climatedata$id))
#unique(climatedata$id)
library(magrittr)
climatedata %<>% separate(id, into=c("gcm", "rcp", "year"), sep="_")
climatedata$year <- as.numeric(climatedata$year)
climatedata <- bind_rows(ewembi, climatedata)

# Climate change of protected areas
clim_pa <- pas %>% left_join(climatedata) %>% 
  filter(`2018` >= 0.5) %>% select(-c(`1819`:`2018`)) %>% 
  gather(variable, val, -c(x,y,gcm,rcp,year)) %>% 
  group_by(x,y, variable, rcp, year) %>% 
  summarise(val=mean(val)) %>% drop_na()
#head(clim_pa)
clim_pa <- filter(clim_pa, variable %in% c("bio1", "bio4", "bio5", "bio12", "bio15", "bio18"))
clim_pa$variable <- factor(clim_pa$variable, 
                           levels=c("bio1", "bio4", "bio5", 
                                    "bio12", "bio15", "bio18"))
clim_pa %>% filter(variable == "bio1") %>% 
  ggplot() + geom_violin(aes(year, val, fill=rcp)) + 
  facet_wrap(~ variable, scales="free", strip.position="left") + 
  theme_bw() + theme(strip.background = element_blank(),
                     strip.placement="outside", 
                     legend.position="bottom") + ylab("") + 
  scale_x_continuous(name="Year", breaks=c(1995,2080),labels=c(1995,2080)) + 
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                    name="Scenario",
                    breaks=c(NA, "rcp26", "rcp60"),
                    labels=c(NA, "RCP2.6", "RCP6.0"))
#ggsave("figures/bio1_pas.png", dpi=600, width=6, height=5)
```

**Fig. 3.** Current and future conditions of bio1 within protected areas.

```{r clim_pas}
clim_pa %>% filter(variable != "bio1") %>% 
  ggplot() + geom_violin(aes(year, val, fill=rcp)) + 
  facet_wrap(~ variable, scales="free", strip.position="left") + 
  theme_bw() + theme(strip.background = element_blank(),
                     strip.placement="outside", 
                     legend.position=c(0.8,0.25)) + ylab("") + 
  scale_x_continuous(name="Year", breaks=c(1995,2080),labels=c(1995,2080)) + 
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                    name="Scenario",
                    breaks=c(NA, "rcp26", "rcp60"),
                    labels=c(NA, "RCP2.6", "RCP6.0"))
#ggsave("figures/clim_pas.png", dpi=600, width=6, height=5)
```

**Fig. 4.** Current and future conditions of bio4, bio5, bio12, bio15, bio18 and bio19 within protected areas.

```{r, echo=F}
# Plot temperature against precipitation

#envdata <- clim_pa %>% filter(variable %in% c("bio1", "bio12")) %>%
#  filter(rcp %in% c("1995", "rcp26")) %>%
#  tidyr::spread(variable, val)

## Calculate minimum convex polygon of climatic niche
# With percent=100, same result as chull function!

# Remove outliers
#library(outliers)
#envdata %<>% ungroup %>% select(year,bio1,bio12)
#envdata <- rm.outlier(envdata, fill=T)

#sp::coordinates(envdata) <- ~bio12+bio1
#mcp1 <- adehabitatHR::mcp(envdata[envdata$year == 1995,], percent=99)
#mcp2 <- adehabitatHR::mcp(envdata[envdata$year == 2080,], percent=99)
#envdata <- data.frame(envdata)
#ggplot() + geom_point(envdata[envdata$year == 1995,], aes(x=bio12,y=bio1), 
#                      fill=NA, colour="black",alpha=0.1) + 
#  geom_point(envdata[envdata$year == 2080,], aes(x=bio12,y=bio1), 
#             fill=NA, colour="red", alpha=0.1) + 
#  geom_polygon(mcp1, aes(x=long, y=lat), colour="black", alpha=0.05, fill=NA) +
#  geom_polygon(mcp2, aes(x=long, y=lat), colour="red", alpha=0.05, fill=NA) + 
#  theme_bw() + labs(x="", y="")
```

```{r sr_pa}
########################################

# Species Data

sr_1995 <- read.csv("data/sr_predicted_ssdm_1995_disp1.csv.xz")
sr_2080 <- read.csv("data/sr_predicted_ssdm_disp1_2080.csv.xz")
sr <- full_join(sr_1995, sr_2080) %>% 
  gather(var, value, -c(x,y, model,group))
sr$var <- gsub("[.]", "-", sr$var)
sr %<>% separate(var, into=c("gcm", "rcp", "year"), sep="_")
sr$year[sr$rcp == "1995"] <- 1995
rm(sr_1995, sr_2080)

########################################

# Combine datasets

climatedata$year <- as.numeric(climatedata$year)
sr$year <- as.numeric(sr$year)
data <- left_join(sr, climatedata)
data <- pas %>% select(x,y,"1995", "2018") %>% left_join(data)
rm(climatedata, pas, sr)

########################################

# Species richness change of protected areas

sr_pa <- data %>% filter(2018 >= 0.5) %>% 
  select(x,y,model,group,gcm,rcp,year,value) %>% 
  group_by(x,y, group, rcp, year) %>% 
  summarise(val=mean(value)) %>% drop_na()
#head(sr_pa)

sr_pa$group <- factor(sr_pa$group, labels=c("Amphibians", "Birds", "Mammals"))
sr_pa %>% ggplot() + geom_violin(aes(year, val, fill=rcp)) + 
  facet_wrap(~ group, scales="free") + ylab("Species richness") + theme_bw() + 
  theme(strip.background = element_blank()) + 
  scale_x_continuous(name="Year", breaks=c(1995,2080), labels=c(1995,2080)) + 
  #scale_y_continuous(limits=c(0,NA), expand=c(0,0)) + 
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                    name="Scenario",
                    breaks=c(NA, "rcp26", "rcp60"),
                    labels=c(NA, "RCP2.6", "RCP6.0"))
#ggsave("figures/sr_pas.png", dpi=600, width=8, height=4)
```

**Fig. 5.** Current (1995) and future (2080) species richness of amphibians, birds and mammals within protected areas.

```{r delta_sr_pas_country, fig.width=10, fig.height=8}
##############################

# Identify the number of species protected by PAs for each country

countries <- readr::read_csv("data/countries_05deg.csv.xz")

library(magrittr)
sr_coun <- sr_pa %>% left_join(countries)
sr_sub <- sr_coun[sr_coun$rcp == 1995,]
sr_sub$rcp[sr_sub$rcp == 1995] <- "rcp26"
sr_coun$rcp[sr_sub$rcp == 1995] <- "rcp60"
sr_coun %>% bind_rows(sr_sub) %>%
  group_by(x,y,year,rcp, NAME) %>%
  summarise(sum=sum(val)) %>% tidyr::spread(year, sum) %>%
  mutate(val=`2080`-`1995`) %>%
  ggplot() + 
  geom_boxplot(aes(as.factor(NAME), val, fill=rcp),
               position = position_dodge2(preserve = "single")) + 
  labs(x="Country", y="Change in SR") + coord_flip() + 
  theme_bw() + theme(strip.background = element_blank()) + 
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                    name="Scenario",
                    breaks=c(NA, "rcp26", "rcp60"),
                    labels=c(NA, "RCP2.6", "RCP6.0"))
#ggsave("figures/delta_sr_pas_country.png", dpi=600, width=5, height=10)
```

**Fig. 6.** Change in total species richness (2080-1995) within protected areas for each country.

```{r delta_sr_pas_continents}
##############################

# Identify the number of species protected by PAs for each continent

continents <- readr::read_csv("data/continents_05deg.csv.xz")

library(magrittr)
sr_con <- sr_pa %>% left_join(continents)
sr_sub <- sr_con[sr_con$rcp == 1995,]
sr_sub$rcp[sr_sub$rcp == 1995] <- "rcp26"
sr_con$rcp[sr_con$rcp == 1995] <- "rcp60"
sr_con %>% bind_rows(sr_sub) %>%
  group_by(x,y,year,rcp, continent) %>%
  summarise(sum=sum(val)) %>% tidyr::spread(year, sum) %>%
  mutate(val=`2080`-`1995`) %>%
  ggplot() + 
  geom_violin(aes(as.factor(continent), val, fill=rcp)) + 
  labs(x="Continent", y="Change in SR") + coord_flip() + 
  theme_bw() + theme(strip.background = element_blank()) + 
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                    name="Scenario",
                    breaks=c(NA, "rcp26", "rcp60"),
                    labels=c(NA, "RCP2.6", "RCP6.0"))
```

**Fig. 7.** Change in total species richness (2080-1995) within protected areas of each continent.

```{r delta_sr_pas_realms}
##############################

# Identify the number of species protected by PAs for each continent

realms <- readr::read_csv("data/realms_05deg.csv.xz")

library(magrittr)
sr_realm <- sr_pa %>% left_join(realms)
sr_sub <- sr_realm[sr_realm$rcp == 1995,]
sr_sub$rcp[sr_sub$rcp == 1995] <- "rcp26"
sr_realm$rcp[sr_realm$rcp == 1995] <- "rcp60"
sr_realm %>% bind_rows(sr_sub) %>%
  group_by(x,y,year,rcp, realm) %>%
  summarise(sum=sum(val)) %>% tidyr::spread(year, sum) %>%
  mutate(val=`2080`-`1995`) %>%
  ggplot() + 
  geom_violin(aes(as.factor(realm), val, fill=rcp)) + 
  labs(x="Realm", y="Change in SR") + coord_flip() + 
  theme_bw() + theme(strip.background = element_blank()) + 
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                    name="Scenario",
                    breaks=c(NA, "rcp26", "rcp60"),
                    labels=c(NA, "RCP2.6", "RCP6.0"))
```

**Fig. 8.** Change in total species richness (2080-1995) within protected areas of each realm.

```{r delta_sr_pas_biomes}
##############################

# Identify the number of species protected by PAs for each continent

biomes <- readr::read_csv("data/biomes_05deg.csv.xz")

library(magrittr)
sr_bio <- sr_pa %>% left_join(biomes)
sr_sub <- sr_bio[sr_bio$rcp == 1995,]
sr_sub$rcp[sr_sub$rcp == 1995] <- "rcp26"
sr_bio$rcp[sr_bio$rcp == 1995] <- "rcp60"
sr_bio %>% bind_rows(sr_sub) %>%
  group_by(x,y,year,rcp, biome) %>%
  summarise(sum=sum(val)) %>% tidyr::spread(year, sum) %>%
  mutate(val=`2080`-`1995`) %>%
  ggplot() + 
  geom_violin(aes(as.factor(biome), val, fill=rcp)) + 
  labs(x="Realm", y="Change in SR") + coord_flip() + 
  theme_bw() + theme(strip.background = element_blank()) + 
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                    name="Scenario",
                    breaks=c(NA, "rcp26", "rcp60"),
                    labels=c(NA, "RCP2.6", "RCP6.0"))
```

**Fig. 9.** Change in total species richness (2080-1995) within protected areas of each biome.
